Type Evolution in Interactive Programming Systems

### 2024-05-31 paper.tex
Tweaks in metamodel, conclusions, etc

% \definecolor{progblue}{RGB}{64,104,224}
% \renewenvironment{leftbar}[1][\hsize]
% {%
%     \def\FrameCommand
%     {%
%         {\color{progblue}\vrule width 3pt}%
%         \hspace{0pt}%must no space.
%         \fboxsep=\FrameSep\colorbox{white}%
%     }%
%     \MakeFramed{\hsize#1\advance\hsize-\width\FrameRestore}%
% }
% {\endMakeFramed}

@@

\author[a]{Jonathan Edwards}%[0000-0003-1958-7967]
\authorinfo{is TBD. Contact him at \email{jonathanmedwards@gmail.com}.}
\affiliation[a]{Independent, Boston, MA, USA}

\author[b]{Tomas Petricek}%[0000-0002-7242-2208]
\authorinfo{is TBD. Contact him at \email{tomas@tomasp.net}.}
\affiliation[b]{Charles University, Prague, Czechia}

\author[c,d]{Tijs van der Storm}%[0000-0001-8853-7934]
\authorinfo{is TBD. Contact him at \email{storm@cwi.nl}.}
\affiliation[c]{Centrum Wiskunde \&\ Informatica (CWI), Amsterdam, Netherlands}
\affiliation[d]{University of Groningen, Groningen, Netherlands}

\author[e]{Geoffrey Litt}%[0000-0003-0858-5165]
\authorinfo{is TBD. Contact him at \email{gklitt@gmail.com}.}
\affiliation[e]{Ink \& Switch, Planet Earth}

@@

Many past improvements in the practice of programming worked by speeding up feedback loops,
both in collaboration with other programmers (through collaborative development tools
\cite{ProGit, goldman2011real,kurniawan2015coder,replit}) and in individual's interaction with the
programming environment (through live programming \cite{rein2018exploratory}).

To make programming more effective, future programming systems will need to further tighten
the feedback loop. They will need to enable programmers to collaborate at a fine-grained level,
letting them adopt individual changes created by their collaborators (as in local-first software
\cite{localfirst}). To provide live feedback, future programming systems will also need to
transcend treating state as ephemeral and reset when the
program is edited and rerun.

To conceive of such systems, we need to shift our focus to \emph{programming systems} \cite{techdims}
where program state is persistent and co-evolves with the behavior and data of the system. This is
already the case in collaborative applications with state persistent in a database, image-based
systems like Smalltalk and live programming environments with hot-reloading where the program can
be edited without resetting all state.

While we are firmly convinced that more collaborative and live programming systems are the
future of programming, there are a number of problems that need to be resolved in order for
such systems to become a reality. A key problem is \emph{type evolution}, i.e., how to
gracefully adapt the shape of data used in a program while preserving integrity
constraints, existing data and the logic of code.

Type evolution is not a new problem. It exists in various forms in a range of existing systems.
The problem of schema evolution in databases is well-studied \cite{erhard06}.
It typically considers data integrity, although co-evolution of database schema
and code has also been studied \cite{Cleve2006,wang19}.
In live programming, the problem is known as hot code reloading \cite{barenz2020essence,beckmann2021shortening},
although it typically focuses on updating transient state and ignores the problem of more complex
data migrations. The problem is also studied in the bidirectional transformation (BX)
community \cite{czarnecki2009bidirectional}, in work on model-driven development
\cite{Cicchetti11,alanen2003} and in more low-level work on dynamic software updating
\cite{hicks2005dynamic}.

Looking at type evolution from the perspective of programming systems can offer a new perspective that
unfies the many different aspects of the problem partially explored in other areas. A programming
system may have permanent data (as in a database) whose structure can change, requiring
corresponding change to data and code. A programming system may have a transient state conforming
to a type structure that may change during live programming. A programming system may also need
to allow programmers to independently work on different versions of a program.

In this paper, we draw connections between these different kinds of type evolution, pointing out
an essential shared structure. First, we present a shared conceptual
framework for thinking about the relationship between schema, code, and data in a programming
system. Second, we present five case studies that use the framework to study type evolution in a
highly diverse set of real-world programming scenarios. We hope this will encourage knowledge
sharing across different subfields, motivate further research on this important problem and
offer inspiring challenges that developers of future live and collaborative programming systems
will attempt to tackle.

@@

\section{Entity Evolution in Data-Centric Systems}
If we see programming systems as stateful environments where programs co-exist with their persistent
data, we can gain valuable insights from research on database systems.
Although schema evolution is well-studied in databases (Section~\ref{sec:related}), past work
does not resolve all problems in our second challenge.

@@

\frameworkbox{
\begin{wrapfigure}{r}{14em}
\vspace{-1.5em}
\includegraphics[width=14em]{figures/arr-backport.png}
\vspace{-2.5em}
\end{wrapfigure}
\frameworkboxtitle{Bidirectional Data Migration}
The scenario adds the need for transferring data changes from a new type to an older type.
In the diagram on the right, we only show data and type, ignoring the code added by the
accounting department. We start with data $D_1$ of type~$T_1$. The accounting deparment
evolves the type (via Extract Entity), obtaining data $D_2$ of type $T_2$ using an automatically
generated forward transformation. It further modifies the data, obtiaing $D_3$. When the
accounting department receives data $D_3$, it needs to migrate it back from type $T_2$ to
$T_1$ using an automatically generated backward transformation. In other words,
type evolution needs to be accompanied with both forward and backward projection for data
(and code) migration.}

@@

\begin{figure}[t]
\centering
\begin{minipage}{0.3\textwidth}
\includegraphics[width=\textwidth]{figures/metamodel.pdf}
\end{minipage}
\hspace*{2pt}
\vline
\begin{minipage}{0.3\textwidth}
  \includegraphics[width=\textwidth]{figures/doors-ast.pdf}
  \end{minipage}
\hspace*{2pt}
\vline
\begin{minipage}{0.3\textwidth}
  \begin{lstlisting}[language=java,morekeywords={machine,on,state,var},basicstyle=\footnotesize,numbers=none]
machine Doors
var isClosed: bool = true
state closed
  isClosed := true
  on open => opened
state opened
  on close => closed
\end{lstlisting}
\end{minipage}
\caption{Definition of the abstract syntax state machine language as a UML metamodel and an example instance, as object diagram and as source code}
\label{FIG:statemachineUML}
\end{figure}

@@

We considered live programming in the context of the Elm architecture in Section~\ref{sec:elm},
where the challenge was to migrate run-time state (data) to a new type to keep the program running.
Work on language modeling offers a new perspective on the problem. In the context of executable domain-specific modeling languages~\cite{}, one can see a program in a domain-specific language as an instance of a metamodel (a static type) that
defines the AST structure of the domain-specific language. The program in a domain-specific language
then in turn determines (defines/implies/induces) a run-time schema: the run-time structures
of code (e.g., classes, inheritance links, declarations, method definitions etc.), as well as the structure of the run-time state (the heap, slots for user data, etc.).
Whereas the example discussed in this section is framed in terms of metamodels and models, this should be seen as an expository device: the principles and challenges apply, e.g., in more grammar-oriented settings.

In the absence of live programming, the run-time \emph{code} structures do not typically
change while using an application and the run-time \emph{state} changes constantly.
Live programming, however, requires reconciling changes to the program with the induced run-time structures
of both the code and the state. This typically means that at a certain point (a quiescent point)
during execution, the code structures need to be updated (hot swapped), and the state needs to be \textit{migrated}.

\subsection{Example}
Consider the example of a simple state machine language with on-entry actions, and typed global
variables (loosely inspired by the SML language of \citet{vanRozen19}). 
The abstract syntax (as UML class diagram) is shown on the left of Figure~\ref{FIG:statemachineUML}. 
The middle of the figure shows an instance: a state machine modeling the opening and closing of a door (in UML object notation). 
The right shows a possible textual concrete syntax modeling the same state machine. 
The state machine has one global
boolean variable, \lstinline{isClosed}, which is flipped according to transitioning between the
\lstinline{closed} and \lstinline{opened} states. For the sake of brevity, the abstract syntax
model  shown on the left of the figure
omits the details for the on-entry actions (\lstinline{Stmt}) and variable types (\lstinline{Type}).

@@

In the following we assume that the state machine is executed by an interpreter that processes incoming events, and traverses run-time pointers representing the transitions accordingly, executing on-entry actions, where the notion of the current state and the values of the global variables are the collective run-time state of the program. The structure that the interpreter operates on can be described as a schema as well, but derived from the abstract syntax scheme shown on the left of Figure~\ref{FIG:statemachineUML}, but decorated with additional properties and associations modeling the run-time state required for execution. This is illustrated in Figure~\ref{FIG:doorsRuntime}. The left shows an augmented metamodel adding properties \lstinline{vars} (to model variable assigments) and \lstinline{visited} (counting number of state visits) to the classes \lstinline{Machine} and \lstinline{State}, respectively.

@@

\frameworkbox{
\begin{wrapfigure}{r}{14.5em}
\vspace{-0.5em}
\includegraphics[width=14.5em]{figures/arr-model.png}
\vspace{-1.5em}
\end{wrapfigure}
\frameworkboxtitle{TODO}
Here, data $D_0$ now represent the state machine definition.
Live data that we obtain by instantiating the state machine still contain this state machine
definition, but also the necessary run-time state required for execution.

This is used to synthesize live machine state $+D_0$ conforming to live machine type $+T_0$.
This then runs producing $+T'_0$ and so on...

If the state machine definition changes from $D_0$ to $D_1$,
we can synthesize a new live machine state $+D_1$ conforming to new live machine type $+T_1$.

The challenge is to merge these changes with the currently running state machine $+T'_0$.
We have to produce merged live machine type $+T_0 \diamond +T_1$ and also merge the
live state changes to obtain $+D'_0 \diamond +D_1$ of the new merged live machine type.
}

\begin{figure}[t]
  \centering
\begin{minipage}{0.5\textwidth}
\includegraphics[width=\textwidth]{figures/runtime-model.pdf}
\end{minipage}
\hspace*{2pt}
\vline
\begin{minipage}{0.4\textwidth}  
\includegraphics[width=\textwidth]{figures/doorsmachine.pdf}
\end{minipage}
\caption{Augmented run-time metamodel for state machines and a run-time instance of the Doors state machine during execution.}
\label{FIG:doorsRuntime}
\end{figure}

An instance of a running state machine consists of an object graph conforming to the run-time schema of state machines, updating the \lstinline{current} state pointer in response to events. This object graph is graphically depicted in Figure~\ref{FIG:doorsRuntime} as a UML object diagram, excluding the statement objects representing on-entry actions. As the diagram shows, the running machine has a current state (\lstinline{closed}) and the \lstinline{isClosed} field has an actual value (\lstinline{true}) in the \lstinline{vars} map. It is instructive to note that the diagram encodes both static and dynamic aspects of the state machine language. Live editing the state machine then requires to \textit{patch}~\cite{SemanticDeltas} this run-time structure without shutting it down, possibly requiring migration of run-time state.

\subsection{Challenge}

\begin{table}[t]
  \centering
\begin{tabular}{lll}\toprule
\textsc{Change} & \textsc{What} & \textsc{How}\\\midrule
add & state & simple  \\
remove & state & structurally simple, but heuristic if state is current \\
rename & state & simple\\
add & variable & add entry to map and initialize \\
remove & variable & remove from the variable map (NB: assumes variable is unused)\\
rename & variable & rename in class, preserving value\\
type change & variable & migrate value in map \\
add & transition & simple \\
remove & transition & structurally simple, but heuristic for pending events\\
event change & transition & structurally simple, but heuristic for pending events\\
add/remove & statement & hotswap code at quiescent point of interpreter\\
\bottomrule
\end{tabular}
\caption{Possible program changes and how to deal with them at run time}
\label{tbl:smchanges}
\end{table}

The space of possible (well-formed\footnote{The abstract syntax schema is not expressive enough to define all static invariants of the language, such as: states must by unique by name, references to variables in actions must be declared, etc.}) changes to a state machine are summarized in Table~\ref{tbl:smchanges}.
The first column indicates the change category, the second the affected kind of object, and third a short description of how to deal with the respective change category. Some notes about this table:
\begin{itemize}
\item all rows with ``simple'' in the third column only require a quiescent~\cite{Tranquility} point in the interpreter loop to update the run-time structure (e.g., as shown in Figure~\ref{FIG:doorsRuntime}).
\item removing a state, however, is only \textit{structurally} simple, since removal is easy, but special care is required if the subject of removal is the current state. In this case, some heuristic is needed, such as: reject the edit, point the current state to the initial state, or some other strategy (e.g., the nearest state, previous state etc.)
% \item everywhere a row mentions ``migrate class'' in the third column, data migration is needed: the run-time objects conforming to the old class must be transformed to instances of the updated class, similar to how Smalltalk migrates objects using \lstinline{become:}.
\item removing a transition or changing the event of a transition potentially has to deal with pending events (e.g., in an event queue) expecting such transitions, since such events are now potentially stale. Strategies to deal with this situation include: simply dropping the events, or requiring that such edits cannot be patched at run time when there are pending events.
\item the type change edit of variable requires a strategy for the current value: either discard and reinitialize, or perform value conversion. For instance, if the type change is from boolean to integer, then true could become 1, and false could become 0.
\item note finally that rename variable is mentioned explicitly as a change: this makes it possible to preserve the run-time value of the variable.
\end{itemize}

\subsection{Goals}

The goals for this challenge are twofold: from the end-user perspective and from the language engineering perspective. A particular challenge from the end-user perspective is to ``do minimal harm'': it is essential for fluid programming experience that the automatically triggered migrations are in a sense as near as possible to the previous application state, to not surprise or confuse the user/developer.
% One approach  considered in earlier work~\cite{RuntimeConstraint} employs the constraint solver Z3 to find a ``nearest'' run-time instance compatible with a source change. Nevertheless, the patching of the run-time state should be quick enough so as to not disrupt the programmer experience.

From the language engineering perspective the goal is to employ techniques, formalisms, and tools, to make the construction of such languages easier. The above example state machine DSL is derived from earlier work~\cite{vanRozen19}, where the authors manually implemented the run-time patch operation and concluded that even for such a simple language (simpler even than the example above) it is a complex and error-pone endeavor. Furthermore, the field of software language engineering studies and develops generic and reusable techniques to improve the development of DSLs and programming languages, for instance, in the context of language workbenches~\cite{ERDWEG201524}. The development of live programming languages, however, is currently still out of reach for all existing language workbenches.

The above case study is concerned with changes to the DSL program (e.g., the Doors statemachine). However, there is another level of schema change to be considered: what if the metamodel (cf. Figure~\ref{FIG:statemachineUML}) itself is edited? In this case both the programs (e.g., Doors) possibly need to be migrated (in case of a change that is not backwards compatible), \textit{and} the running instance, including the interpreter, since its structure is, too, governed by the abstract syntax of the language.

@@

\subsection{Extensions}

While the above example is arguably simple, the problem becomes much more challenging when the programs themselves define data types, classes, records, structures etc. Since possibly many instances (values, objects) of such data types may exist at run-time, these all have to be migrated in such a way that programmer experience is minimally disrupted, and that the invariants of said data types is maintained.
Another extension, tying in with the data-oriented examples above, involves refactorings of data types in a program. Typically, a refactoring should be behavior preserving, but can it also preserve run-time data? The minimal example is the consistent variable rename in Table~\ref{tbl:smchanges}, which should not have any effect on run-time state.  A more complex refactoring is described in Section~\ref{sec:elm}.

@@

\section{\textcolor{red}{[TVDS] Super rough attempt at the paragraph}}

Modern day software development is collaborative. Single user programming does not exist anymore.
Just as document editing is becoming more collaborative through online systems like pioneered by Google Docs,
programming would benefit form online collaboration. In a sense this is live programming++: whereas traditional
live programming systems assume a single developer editing code and see their changes immediately reflected
in the running system, online collaborative programming is the same but for multiple programmers at the same time.
The key challenge here, we posit, the *schema change*. We take inspiration from database migration,
model evolution, and version control, and identify benchmark problems. This is already evident in live programming:
editing a program, consisting of type definitions defining the structure run-time state and (transient) procedures, functions
etc. operating on that state, requires run-time state to be migrated, after the programmer edits their program.
Adding the dimension of multiple programs brings us into the domain local-first, and decentralized version control, where
copies are created and edited, which eventually have to be reconciled again (e.g., through merging).
Whereas schema evolution and its associated migration problems has seen a lot of attention in the database world,
the model-driven world, and the bidirectional transformation (BX) communities, the techniques are still too low-level,
too manual, and too much focused on a single user perspectives. By identifying the similarities between analogous
problems in live, multi-user programming systems, we hope to stimulate further research on more comprehensive, automatic,
high-level, and online solutions. So that the benefits of live programming can be enjoyed in a global and distributed setting
involving multiple programmers at the same time.

\section{\textcolor{red}{[JE] Type evolution for live programming}}

Many past improvements in the practice of programming worked by speeding up feedback loops. Live Programming aspires to the optimum of immediate feedback. The problem is that modern software is built out of many layers of different technologies -- thoroughly live programming would propagate changes immediately up and down the entire stack of a running system, from the database to the code, and from the client to the server. We call this \textit{full-stack} live programming. Another problem is that modern software is built with complex collaborations involving development, testing, deployment, and suppport. Such workflows are themselves feedback loops -- thoroughly live programming would propagate changes between participants frictionlessly. We call this \textit{end-to-end} live programming.

We propose a research agenda towards this greater vision of live programming. The starting point is database \textit{schema evolution}, which propagates database schema changes into data migrations and query rewriting. We call for generalized \textit{type evolution} that works immediately across the entire software stack, and automatically across the entire software workflow. This paper contributes a set of challenge problems to serve as targets for research on type evolution and as a basis of comparison for the results.

\section{\textcolor{red}{[GL] Rough paragraph}}

Most programming languages analysis treats state as ephemeral: tied to the lifetime of a running program, and reset when the program is edited. But when we take a broader view, most \textit{programming systems} (cite) have state that lasts beyond the lifetime of a single execution. This includes collaboration applications with state persisted in a database, image-based systems like Smalltalk, and live programming environments where the program can be edited without resetting all state.

Although these programming systems exist in very different contexts, they all encounter similar problems related to the essential difficulty of preserving state across program edits. New versions of the program must correctly handle pre-existing data. When the data schema or type definitions change, existing data must be evolved to conform to the new schema. Changing data representations or types can also require changing corresponding parts of the program. These problems have been noticed, named, and studied in these different programming systems: ``schema evolution'' in the databases literature, ``hot reloading'' in live programming, etc.

In this paper, we draw connections between these different contexts, pointing out essential shared structure in the hopes of motivating further research on this important problem and encouraging knowledge sharing across subfields. Our contributions are: 1) a shared framework for thinking about the relationship between schema, code, and data in a programming system, 2) Case studies of the framework in a highly diverse set of programming systems, motivated by real-world examples.

\section{\textcolor{red}{[TP] Rough paragraph}}

Programming is increasingly collaborative and interactive. We welcome this trend, because
greater collaboration and interactivity make programming more effective. Future programming
environments enable programmers to collaborate at a more fine-grained level, adopting individual
code changes created by their collaborators as needed. They also let programmers interactively
edit code, often during program execution.

In order to support this new way of working, programming systems increasingly have to tackle the
problem of schema evolution. If a programmer imports a change that edits a schema, code and data
that relies on such schema needs to be adapted to match the new schema. Similarly, if a programmer
in a live programming environment modifies their type definition, current program state needs to
be updated to match the new structure of types. This is a well-understood and well-studied problem
in database community, but it has received little attention in work on programming systems.

>>>>

@@

\author[a]{Jonathan Edwards}%[0000-0003-1958-7967]
\authorinfo{is an independent researcher. Contact him at \email{jonathanmedwards@gmail.com}.}
\affiliation[a]{Independent, Boston, MA, USA}

\author[b]{Tomas Petricek}%[0000-0002-7242-2208]
\authorinfo{is an Associate Professor at Charles University. Contact him at \email{tomas@tomasp.net}.}
\affiliation[b]{Charles University, Prague, Czechia}

\author[c,d]{Tijs van der Storm}%[0000-0001-8853-7934]
\authorinfo{is a Senior Researcher at Centrum Wiskunde \&\ Informatica (CWI) and a Professor at the University of Groningen. Contact him at \email{storm@cwi.nl}.}
\affiliation[c]{Centrum Wiskunde \&\ Informatica (CWI), Amsterdam, Netherlands}
\affiliation[d]{University of Groningen, Groningen, Netherlands}

\author[e]{Geoffrey Litt}%[0000-0003-0858-5165]
\authorinfo{is a Senior Researcher at the independent research lab Ink \& Switch. Contact him at \email{gklitt@gmail.com}.}
\affiliation[e]{Ink \& Switch, Planet Earth}

@@

Many past improvements in the practice of programming worked by speeding up feedback loops,
both in collaboration with other programmers (via collaborative development tools
\cite{ProGit, goldman2011real,kurniawan2015coder,replit}) and in individual's interaction with the
programming environment (through live programming \cite{rein2018exploratory}).
%
To make programming more effective, future programming systems will need to further tighten
the feedback loop. They will need to enable programmers to collaborate at a fine-grained level,
letting them adopt individual changes created by their collaborators (as in local-first software
\cite{localfirst}). To provide live feedback, future programming systems will also need to
transcend treating state as ephemeral and reset when the
program is edited and rerun.

To conceive of such systems, we need to shift our focus to \emph{programming systems} \cite{techdims}
where program state is persistent and co-evolves with the behavior and data of the system. This view
is already used in collaborative applications with state persistent in a database, image-based
systems like Smalltalk and live programming environments with hot-reloading where the program can
be edited without resetting all state.

While we are firmly convinced that more collaborative and live programming systems are the
future of programming, there are a number of problems that need to be resolved in order for
such systems to become a reality. A key problem is \emph{type evolution}: how to
gracefully adapt the shape of data used in a program while preserving integrity
constraints, existing data and the logic of code.

Type evolution is not a new problem. It exists in various forms in a range of existing systems.
The problem of schema evolution in databases is well-studied \cite{erhard06}.
It typically considers data integrity, although co-evolution of database schema
and code has also been studied \cite{Cleve2006,wang19}.
In live programming, the problem is known as hot code reloading \cite{barenz2020essence,beckmann2021shortening},
although it typically focuses on updating transient state and ignores the problem of more complex
data migrations. The problem is also studied in the bidirectional transformation (BX)
community \cite{czarnecki2009bidirectional}, in work on model-driven development
\cite{Cicchetti11,alanen2003} and in more low-level work on dynamic software updating
\cite{hicks2005dynamic}. We draw on extensive literature review, included in Appendix~\ref{sec:related}.

Looking at type evolution from the perspective of programming systems can offer a new perspective that
unfies the many different aspects of the problem partially explored in other sub-fields. A programming
system may have permanent data (as in a database) whose structure can change, requiring
corresponding change to data and code. A programming system may have a transient state conforming
to a type structure that may change during live programming. A programming system may also need
to allow programmers to independently work on different versions of a program.

In this paper, we draw connections between these different kinds of type evolution, pointing out
an essential shared structure. First, we present a shared conceptual
framework for thinking about the relationship between schema, code, and data in a programming
system. Second, we present six case studies that use the framework to study type evolution in a
highly diverse set of real-world programming scenarios. We hope this will encourage knowledge
sharing across different subfields, motivate further research on this important problem and
offer inspiring challenges that developers of future live and collaborative programming systems
will attempt to tackle.

@@

\section{Entity Evolution in Data-Centric Systems}
If we see programming systems as stateful environments where programs co-exist with their persistent
data, we can gain valuable insights from research on database systems.
Although schema evolution is well-studied in databases (Appendix~\ref{sec:related}), past work
does not resolve all problems in our second challenge.

@@

\frameworkbox{
\begin{wrapfigure}{r}{14em}
\vspace{-1.5em}
\includegraphics[width=14em]{figures/arr-backport.png}
\vspace{-2.5em}
\end{wrapfigure}
\frameworkboxtitle{Bidirectional Data Migration}
The scenario adds the need for transferring data changes from a new type to an older type.
In the diagram on the right, we only show data and type, ignoring the code added by the
accounting department. We start with data $D_1$ of type~$T_1$. The accounting deparment
evolves the type (via Extract Entity), obtaining data $D_2$ of type $T_2$ using an automatically
generated forward transformation. It further modifies the data, obtiaing $D_3$. When the
accounting department receives data $D_3$, it needs to migrate it back from type $T_2$ to
$T_1$ using an automatically generated backward transformation. That is,
type evolution needs to be accompanied with both forward and backward projection for data
(and code) migration.}

@@

\begin{figure}[t]
\begin{adjustwidth}{-3em}{-3em}
\begin{minipage}{0.4\textwidth}
\includegraphics[width=\textwidth]{figures/metamodel.pdf}
\end{minipage}
\hspace*{2pt}
\vline
\begin{minipage}{0.4\textwidth}
  \includegraphics[width=\textwidth]{figures/doors-ast.pdf}
  \end{minipage}
\hspace*{2pt}
\vline
\begin{minipage}{0.3\textwidth}
  \begin{lstlisting}[language=java,morekeywords={machine,on,state,var},basicstyle=\footnotesize,numbers=none]
machine Doors
var isClosed: bool = true

state closed
  isClosed := true
  on open => opened

state opened
  isClosed := false
  on close => closed
\end{lstlisting}
\end{minipage}
\end{adjustwidth}
\caption{Definition of the abstract syntax state machine language as a UML metamodel (left) and
an example instance, as object diagram (middle) and as code (right).}
\label{FIG:statemachineUML}
\end{figure}

@@

We considered live programming in the context of the Elm architecture in Section~\ref{sec:elm},
where the challenge was to migrate run-time state (data) to a new type to keep the program running.
Work on language modeling offers a new perspective on the problem.
In the context of executable domain-specific modeling languages~\cite{klint2011easy}, one can see a
program in a domain-specific language as an instance of a metamodel (a static type) that
defines the AST structure of the domain-specific language.

A program in a domain-specific language in turn determines (defines/implies/induces) a run-time
schema: the run-time structures of code (e.g., classes, inheritance links, declarations, method
definitions etc.), as well as the structure of the run-time state (the heap, slots for user data, etc.).
Whereas the example discussed in this section is framed in terms of metamodels and models, this can
be seen as an expository device: the principles and challenges apply in more grammar-oriented settings.
However, the framing also suggests interesting additional challenges.

In the absence of live programming, the run-time \emph{types} and \emph{code} do not change while
running a program in a given domain-specific language and the run-time \emph{state} changes constantly.
Live programming, however, requires reconciling changes to the program with the induced run-time structures
of both the code and the state. This typically means that at a certain point (a quiescent point)
during execution, the code structures need to be updated (hot swapped), and the state needs to be
\textit{migrated}.

Figure~\ref{FIG:statemachineUML} defines a simple state machine domain-specific language with
on-entry actions, and typed global variables (loosely inspired by SML \cite{vanRozen19}).
The figure shows the abstract syntax as UML class diagram (left); an instance modelling the
opening and closing of a door in UML object notation (middle) and a possible textual concrete
syntax modeling the same state machine (right).
%
The state machine has one global Boolean variable, \lstinline{isClosed}, which is flipped when
transitioning between the \lstinline{closed} and \lstinline{opened} states. For the sake of brevity,
the abstract syntax model shown on the left omits the details for the on-entry
actions (\lstinline{Stmt}) and variable types (\lstinline{Type}).

@@

In the following we assume that the state machine is executed by an interpreter that processes
incoming events, traverses the transitions accordingly, and executes on-entry actions. The notion
of the current state and the values of the global variables are the collective run-time state of
the program. The structure that the interpreter operates on can be described by a run-time schema.
This schema (Figure~\ref{FIG:doorsRuntime}, left) is derived from the abstract syntax
schema (Figure~\ref{FIG:statemachineUML}, left), but is decorated with additional properties and
associations modeling the run-time state required for execution. The augmented metamodel includes
properties \lstinline{vars} (to model variable assigments) and \lstinline{visited} (counting
number of state visits) to the classes \lstinline{Machine} and \lstinline{State}, respectively.

@@

\begin{figure}[t]
\vspace{-0.5em}
\centering
\begin{minipage}{0.5\textwidth}
\includegraphics[width=\textwidth]{figures/runtime-model.pdf}
\end{minipage}
\hspace*{2pt}
\vline
\begin{minipage}{0.4\textwidth}
\includegraphics[width=\textwidth]{figures/doorsmachine.pdf}
\end{minipage}
\caption{Augmented run-time metamodel for state machines (left) and a run-time instance
of the Doors state machine during execution (right).}
\label{FIG:doorsRuntime}
\vspace{-0.5em}
\end{figure}

An instance of a running state machine consists of an object graph conforming to the run-time
schema of a state machine, updating the \lstinline{current} state pointer in response to events.
This object graph is depicted as a UML object diagram in Figure~\ref{FIG:doorsRuntime} (right),
excluding the statement objects representing on-entry actions. As the diagram shows, the running
machine has a current state (\lstinline{closed}) and the \lstinline{isClosed} field has a
value (\lstinline{true}) in the \lstinline{vars} map. It is instructive to note that the
diagram encodes both static and dynamic aspects of the state machine language.

\challenge{Live Editing Domain-Specific Languages}
The challenge is editing the state machine (Figure~\ref{FIG:statemachineUML}, right)
created in the domain-specific language. This requires to \textit{patch}~\cite{SemanticDeltas} the
run-time structure (Figure~\ref{FIG:doorsRuntime}, right) without discarding and recreating it,
possibly requiring migration of run-time state.

The space of possible %(well-formed)\footnote{The abstract syntax schema is not expressive enough
%to define all static invariants of the language, such as: states must by unique by name, references
%to variables in actions must be declared, etc.})
changes to a state machine is summarized in Table~\ref{tbl:smchanges}.
The first column indicates the change category, the second the affected kind of object, and third
a description of how to deal with the respective change category. In summary:
\begin{itemize}
\item All rows with ``simple'' in the third column only require a quiescent~\cite{Tranquility} point in the interpreter loop to update the run-time structure (e.g., as shown in Figure~\ref{FIG:doorsRuntime}).
\item Removing a state, however, is only \textit{structurally} simple, since removal is easy, but special care is required if the subject of removal is the current state. In this case, heuristic or user input is needed, such as: reject the edit, point the current state to the initial state, or some other strategy (e.g., the nearest state, previous state etc.)
% \item everywhere a row mentions ``migrate class'' in the third column, data migration is needed: the run-time objects conforming to the old class must be transformed to instances of the updated class, similar to how Smalltalk migrates objects using \lstinline{become:}.
\item Removing a transition or changing the event of a transition potentially has to deal with pending events (e.g., in an event queue) expecting such transitions, since such events are now potentially stale. Strategies to deal with this situation include dropping the events, or preventing the edit when there are pending events.
\item Changing the type of a variable requires a strategy for the current value: either discard and reinitialize, or perform value conversion.
\item Note finally that rename variable is mentioned explicitly as a change: this makes it possible to preserve the run-time value of the variable.
\end{itemize}

\begin{table}[t]
\begin{adjustwidth}{-2em}{-2em}
\begin{tabular}{lll}\toprule
{\firamedium Change} & {\firamedium What} & {\firamedium How}\\\midrule
Add/Rename & state & Simple  \\
Remove & state & Structurally simple, but heuristic if state is current \\
Add & variable & Add entry to map and initialize \\
Remove & variable & Remove from the variable map (NB: assumes variable is unused)\\
Rename & variable & Rename in class, preserving value\\
Type change & variable & Migrate value in map \\
Add & transition & Simple \\
Remove & transition & Structurally simple, but heuristic for pending events\\
Event change & transition & Structurally simple, but heuristic for pending events\\
Add/remove & statement & Hotswap code at quiescent point of interpreter\\
\bottomrule
\end{tabular}
\vspace{0.3em}
\caption{Possible program changes and how to deal with them at run time}
\label{tbl:smchanges}
\end{adjustwidth}
\vspace{-1em}
\end{table}

\frameworkbox{
\begin{wrapfigure}{r}{14em}
\vspace{-0.75em}
\includegraphics[width=14.5em]{figures/arr-model.png}
\vspace{-1.5em}
\end{wrapfigure}
\frameworkboxtitle{Code as Data Migrations}
We could see this challenge as type, code and data co-evolution as in the case of the Elm
architecture in Section~\ref{sec:elm}, but a more interesting perspective is to see the
state machine definition as data $D_0$ conforming to a type $T$ defined by the metamodel.

When the state machine is instantiated, the metamodel and the definition are used to derive
the run-time type ${}^+\!T_0$. The run-time data ${}^+\!D_0$ conforming to this type still
contain the state machine definition, but also the necessary run-time state required for
its execution, which will produce new versions of the run-time data such as ${}^+\!D_0'$.

If the state machine definition changes from $D_0$ to $D_1$ (top right), we can synthesize
a new run-time state machine type ${}^+\!T_1$ and new data ${}^+\!D_1$ conforming to it.
The challenge is to merge the two run-time types instantiated from different definitions
to produce a type ${}^+\!T_0 \diamond {}+\!T_1$, but also merge the run-time state of the
running machine with the new instantiated state to obtain ${}^+\!D'_0 \diamond {}^+\!D_1$.
}

\subsection*{Remarks: Requirements and Extensions}

The goals for this challenge are twofold: from the end-user perspective and from the language engineering perspective. A challenge from the end-user perspective is again to ``do minimal harm'': it is essential for fluid programming experience that the automatically triggered migrations result in a state as near as possible to the previous application state, to not surprise or confuse the user/developer.
% One approach  considered in earlier work~\cite{RuntimeConstraint} employs the constraint solver Z3 to find a ``nearest'' run-time instance compatible with a source change. Nevertheless, the patching of the run-time state should be quick enough so as to not disrupt the programmer experience.

From the language engineering perspective the goal is to employ techniques, formalisms, and tools, to make the construction of such languages easier. The above example is derived from earlier work~\cite{vanRozen19}, where the authors manually implemented the run-time patch operation and concluded that even for such a small language (simpler than the example here) it is a complex error-pone endeavor. Furthermore, the field of software language engineering studies and develops generic and reusable techniques to improve the development of DSLs and programming languages, e.g., in the context of language workbenches~\cite{ERDWEG201524}. The development of live programming languages, however, is currently out of reach for all existing language workbenches.

The above case study is concerned with changes to the DSL program (e.g., the Doors state machine). However, there is another level of schema change to be considered: what if the metamodel (cf. Figure~\ref{FIG:statemachineUML}) itself is edited? In this case both the programs (e.g., Doors) possibly need to be migrated (in case of a change that is not backwards compatible), \textit{and} the running instance, including the interpreter, since its structure is, too, governed by the abstract syntax of the language.

@@

While the above example is arguably simple, the problem becomes more challenging when the programs themselves define data types, classes, records, structures etc. Since possibly many instances (values, objects) of such data types may exist at run-time, these all have to be migrated in such a way that the programmer experience is minimally disrupted, and that the invariants of said data types is maintained.
Another extension, tying in with the data-oriented examples above, involves refactorings of data types in a program. Typically, a refactoring should be behavior preserving, but can it also preserve run-time data? The minimal example is the consistent variable rename in Table~\ref{tbl:smchanges}, which should not have any effect on run-time state.

\section{Conclusions}
We believe that effective future programming systems will need to tighten the feedback loop,
both in making programming more live and in making it more collaborative. One of the key obstacles
to making progress in this direction is the problem of \emph{type evolution}. When the structure
of a program evolves, code and data need to co-evolve to match the new structure.
%
The problem of type evolution has been explored in a wide range of sub-fields, but rarely in its
full complexity. Moreover, different sub-fields often use different language, making knowledge
transfer between them difficult.

In this paper, we developed a conceptual framework
that made it possible to present six case studies from a wide range of areas in a unified way.
The areas included live programming, database schema change, model-driven development,
spreadsheet and document-oriented programming.
%
Our work highlights the importance
of the problem, provides a new uniform way of thinking about it and provides numerous specific
challenges that can be used to evaluate the capabilities of new programming systems that aim
to tackle the problem.

We did not present new solutions to the problem, but we hope to inspired the reader to
do so. Indeed, we look forward to novel solutions to the challenges presented in this paper and
the resulting more live and collaborative future of programming!

@@
